Moisture absorption and drying alter nonadditive litter flammability in a mixed conifer forest

### 2020-04-23 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Start on response letter and intro edits

      \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
    and drying can change flammability rankings under moist compared with dry
    fuels because traits that increase flame spread rate increase water
    absorption. Litter mixtures generally burn with fire behavior more similar
    to that of the most flammable constituent species and this effect increases
    as fuels dry.

@@

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
that similar nonadditive effects may influence litter moisture content and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with one study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}. In mixed conifer forest in
California, where the fire regime is characterized by surface fires fueled by
plant litter, overstory tree composition drives variation in local fire
behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
flammability across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns
\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
mixed conifer forests are based largely on burning dry litter.

We asked if differences in litter moisture absorption and drying rates could
change the relative flammability of litter across species compared with
flammability measurements based on dry fuels. We investigated these litter
moisture dynamics across eight tree species in a temperate fire-prone forest in
California. We determined the maximum water absorption and rate of moisture
loss and examined how these traits vary in relation to specific leaf area and
litter packing density. Experimental burn trials of leaf litter at various
moisture levels determined how time since wetting influenced two axes of
flammability characterized by flame spread rate and total fuel consumed. To
determine the potential magnitude of mixture effects, we tested if the
nonadditive effects on reported in previous studies occur under higher moisture
levels, and whether fuel moisture effects strengthen or moderate any
nonadditivity.

>>>>

      \item \emph{Synthesis:} This study is the first to examine ecologically
    relevant drying rates and their interaction with non additive misture
    effects on flammability. We find that taxon specific patterns of moisture
    absorption and drying change flammability rankings under moist compared
    with dry fuels because traits that increase flame spread rate increase
    water absorption. Litter mixtures generally burn with fire behavior more
    similar to that of the most flammable constituent species and this effect
    increases as fuels dry.

@@

In mixed species communities, species traits may interact such that the
flammability of multi-species mixtures is not the average of the constituent
species. Such nonadditive effects are often synergistic in dry litter fuels
where the flammability of a mixture is driven by the most flammable species
\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
nonadditivity has also been recorded for litter mixtures
\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
that similar nonadditive effects may influence litter moisture content and
flammability of moist fuels. There is some evidence for nonadditive
flammability in moist fuels with one study finding increasing nonadditivity
with moisture \citep{Blauw+Wensink+etal-2015}, but this previous work did not
include estimation of the dry-down process and had an unclear relationship to
fire behavior int he field.. In mixed conifer forest in California, where the
fire regime is characterized by surface fires fueled by plant litter, overstory
tree composition drives variation in local fire behavior
\citep{Schwilk+Caprio-2011} as a result of differential litter flammability
across tree species and as a result of synergistic interactions in
multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
flammability feedbacks may help explain historical fire frequency patterns
\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
mixed conifer forests are based largely on burning dry litter.

This study is the first to examine ecologically relevant drying rates and their
interaction with non additive moisture effects on flammability. We asked if
differences in litter moisture absorption and drying rates could change the
relative flammability of litter across species compared with flammability
measurements based on dry fuels. We investigated these litter moisture dynamics
across eight tree species in a temperate fire-prone forest in California. We
determined the maximum water absorption and rate of moisture loss and examined
how these traits vary in relation to specific leaf area and litter packing
density. Experimental burn trials of leaf litter at various moisture levels
determined how time since wetting influenced two axes of flammability
characterized by flame spread rate and total fuel consumed. To determine the
potential magnitude of mixture effects, we tested if the nonadditive effects on
reported in previous studies occur under higher moisture levels, and whether
fuel moisture effects strengthen or moderate any nonadditivity.
